---
title: "Which variable to summarize in `tbl_summary()`"
code-fold: true
categories: [tbl_summary]
---


Quick way to remember where to put what variable: 

- `by`: stratification, as column. Typically treatment arm.
- `include`: variables to summarize, as rows. 
  - continuous: value in one row, missing in another row.
  - categorical: one row per category, missing in another row.

To stratify by **more than one variable**, you can make use of the **wide format** data.


### Load data and packages

We are going to use the `trial` data from `gtsummary`. 

```{r}
#| eval: true
#| message: false
#| warning: false
library(gtsummary)
library(dplyr)
library(tidyr)
head(trial)
```


# Basic `tbl_summary()` table

When including both numeric and categorical variables, a long table will be created.

```{r}
#| eval: true
tbl_summary(data = trial, include = c(age, marker, grade))
```

We can stratify by treatment arm by using the `by` argument. 


```{r}
#| eval: true

# stratify by treatment arm
tbl_summary(data = trial, include = c(age, marker, grade), by = trt, missing = "no")
```

# Stratify by 2 variables

In the example above, we stratified by one variable: treatment arm. Other variables are summarized as rows. Notice the indentation: it is still 3 variables: age, marker, grade.

What if you want to stratify between the variables presented in the rows? For example, you want to know what is the **age across different levels of stages**. you can do this:

```{r}
# stratify by stage
tbl_summary(data = trial, include = c(age), by = stage, missing = "no")
```

The disadvantage of this is that you cannot stratify by **treatment arm** at the same time. You need to make use of the wide format data.

### Use wide format

You need to keep the ID variable to make sure the data is in wide format, to keep each row uniquely identifiable.
```{r}
trial_mini <- select(trial, trt, age, stage) |> mutate(id = 1:nrow(trial))
# wide format
trial_mini_wide <- trial_mini |>
  pivot_wider(names_from = stage, values_from = age)

# how long format looks like
head(trial_mini)
# stage variable is expanded into 4 variables: T1, T2, T3, T4
head(trial_mini_wide)
```


Now you can summarize all the stage variables together using `include()`. They will be piled together in rows, all about the target numeric variable, **age**. 

```{r}
tbl_summary(
  trial_mini_wide,
  include = c('T1', 'T2', 'T3', 'T4'),
  missing = "no"
)
```

You should name the variables in the wide format to make sure the labels are informative. Use `label` argument, and specify a list.

```{r}
tbl_summary(
  trial_mini_wide,
  by = trt,
  include = c('T1', 'T2', 'T3', 'T4'),
  missing = 'no',
  label = list(
    'T1' = "Stage T1",
    'T2' = "Stage T2",
    'T3' = "Stage T3",
    'T4' = "Stage T4"
  )
)
```

### What statistics? 

You can specify a different statistics than its default. 

```{r}
tbl_summary(
  trial_mini_wide,
  by = trt,
  include = c('T1', 'T2', 'T3', 'T4'),
  missing = 'no',
  label = list(
    'T1' = "Stage T1",
    'T2' = "Stage T2",
    'T3' = "Stage T3",
    'T4' = "Stage T4"
  ),
  statistic = list(all_continuous() ~ "{mean} ({sd})")
)
```


